Inferring the existence and direction of causal associations in the face of measurement error

### 2017-01-02 docs/manuscript.Rmd
updated

fig_nums(
	"shakhplot",
	caption=paste0("Using ", nrow(qtl_orig), " putative associations between DNA methylation and gene expression we used the Steiger test to infer the direction of causality between them. The rightmost bar shows the proportion of associations for each of the two possible causal directions (colour key) assuming no measurement error in either gene expression or DNA methylation levels. The proportions change when we assume different levels of measurement error in gene expression levels (x-axis) or DNA methylation levels (columns of boxes). If there is systematically higher measurement error in one platform it will appear to be less likely to be the causal factor."),
	display=FALSE
)

@@

### MR analysis of gene expression and DNA methylation

We used the Steiger test to infer the direction of causality between DNA methylation levels and gene expression levels. To do this we obtained a list of `r nrow(qtl_orig)` gene expression - DNA methylation associations as reported in Shakhbazov et al [-@Shakhbazov2016]. These were filtered to be located on the same chromosome, have robost correlations after correcting for multiple testing, and to share a SNP that had a robust cis-acting effect on both the DNA methylation probe and the gene expression probe. Using the reported summary data we used the Steiger test of two independent correlations to identify which direction of causality was most likely for each association. We found that the causal direction commonly goes in both directions (`r cfig("shakhplot")`). We then went on to predict the causal directions of the associations for varying levels of possible measurement error. `r cfig("shakhplot")` shows that the qualitative result of which of DNA methylation or gene expression is more commonly the causal factor is very strongly determined by measurement error.

We performed two sample MR [@Pierce2013] for each association in the direction of causality that we estimated to be the correct direction from the Steiger test. We observed that the sign of the MR estimate was always in the same direction as the Pearson correlation coefficient reported by Shakhbazov et al [-@Shakhbazov2016], and that there was a moderate correlation between the magnitude of the causal correlation and the observational Pearson correlation (`r ctab("shakhtab")`).

We also observed that for associations where methylation caused gene expression the causal effect was more likely to be negative than for the associations where gene expression caused methylation (OR = `r shakhtest2$estimate` (95% CI `r shakhtest2$conf.int[1]` - `r shakhtest2$conf.int[2]`)), suggesting that reducing gene expression levels at a controlling CpG typically leads to increased gene expression levels, consistent with expectation [@Bird2002].

@@

Researchers are often confronted with the problem of making causal inferences using a statistical framework on observational data, and using genetic instruments as surrogates for randomisation is a useful approach. MR and the CIT employ genetic instruments in different ways, and unfortunately neither offer solutions that work perfectly in all scenarios. In the face of measurement error it is evident that causal inference drawn through mediation based methods such as the CIT analysis typically fare worse than MR and will be difficult to interpret, and this is a serious problem considering that it is often impossible to estimate the extent of measurement error present in observational data.

In the epidemiological literature issues of measurement error in mediation is relatively well explored [@Fewell2007]. Our analysis extends this to related  methods such as CIT that are widely used in predominantly 'omic data. These methods are indeed liable to the same issues as standard mediation based analysis, and specifically we show that as measurement error in the exposure variable increases, CIT is likely have reduced statistical power, and liable to infer the wrong direction of causality. We also demonstrate that, though seemingly unintuitive, increasing sample size does not resolve the issue, rather it leads to more extreme p-values for the model that predicts the wrong direction of causality.

@@

Mediation based network approaches are very well established [@Lagani2015] and have a number of extensions that make them valuable tools, including for example network construction. But because they are predicated on the basic underlying principles of mediation they are liable to suffer from the same issues of measurement error. Recent advances in MR methodology, for example applying MR to genetical genomics [@Relton2012], multivariate MR [@Burgess2014a], network MR [@Burgess2015] and mediation through MR [@Varbo2015] may offer more robust alternatives for these more complicated problems.

In our simulations we focused on the simple case of a single instrument in a single sample setting, and assumed that pleiotropy (the influence of the instrument on the outcome through a mechanism other than the exposure) was not present. Recent method developments in MR [@Bowden2015] have focused on accounting for the issues that horizontal pleiotropy can introduce, but how they perform in the presence of measurement error remains to be explored. An important advantage that MR confers over most mediation based analysis is that it can be performed in two samples, which can considerably improve power and expand the scope of analysis. However, whether there is a substantive difference in two sample MR versus one sample MR in how measurement error has an effect is not yet fully understood. We have also assumed no measurement error in the genetic instrument, which is not unreasonable given the strict QC protocols that ensure high quality genotype data is available to most studies. We have restricted the scope to only exploring non-differential measurement error and avoided the complications incurred if measurement error in the exposure and outcome is correlated, however our analysis goes beyond many previous explorations of measurement error by assessing the impacts of both imprecision (noise) and linear transformations of the true variable on causal inference.

>>>>

fig_nums(
	"shakhplot",
	caption=paste0("Using ", nrow(qtl_orig), " putative associations between DNA methylation and gene expression we used the Steiger test to infer the direction of causality between them. a) The rightmost bar shows the proportion of associations for each of the two possible causal directions (colour key) assuming no measurement error in either gene expression or DNA methylation levels. The proportions change when we assume different levels of measurement error in gene expression levels (x-axis) or DNA methylation levels (columns of boxes). If there is systematically higher measurement error in one platform it will appear to be less likely to be the causal factor. b) The relationship between the Pearson correlation between DNA methylation and gene expression levels (x-axis) and the causal estimate (scaled to be in standard deviation units, y-axis). c) Distribution of estimated causal effect sizes, stratified into associations inferred to be due to DNA methylation causing expression (blue) and expression causing DNA methylation (red)."),
	display=FALSE
)

@@

### MR analysis of gene expression and DNA methylation levels

We used the Steiger test to infer the direction of causality between DNA methylation levels and gene expression levels. To do this we obtained a list of `r nrow(qtl_orig)` gene expression - DNA methylation associations as reported in Shakhbazov et al [-@Shakhbazov2016]. These were filtered to be located on the same chromosome, have robost correlations after correcting for multiple testing, and to share a SNP that had a robust cis-acting effect on both the DNA methylation probe and the gene expression probe. Using the reported summary data we used the Steiger test of two independent correlations to identify which direction of causality was most likely for each association. We found that the causal direction commonly goes in both directions (`r cfig("shakhplot")`a). We then went on to predict the causal directions of the associations for varying levels of possible measurement error. `r cfig("shakhplot")`a shows that the qualitative result of which of DNA methylation or gene expression is more commonly the causal factor is very strongly determined by measurement error.

We performed two sample MR [@Pierce2013] for each association in the direction of causality that we estimated to be the correct direction from the Steiger test. We observed that the sign of the MR estimate was generally in the same direction as the Pearson correlation coefficient reported by Shakhbazov et al [-@Shakhbazov2016], but with a few exceptions (`r cfig("shakhplot")`b). There was a moderate correlation between the absolute magnitudes of the causal correlation and the observational Pearson correlation (r = `r cor(abs(shakhbazov$mr_r[real_index]), abs(shakhbazov$pearson[real_index]))`). Together these inferences suggest that even for relatively low level associations it is important to use causal inference methods over observational associations to infer causal effect sizes.

We also observed that for associations where methylation caused gene expression the causal effect was more likely to be negative than for the associations where gene expression caused methylation (OR = `r shakhtest2$estimate` (95% CI `r shakhtest2$conf.int[1]` - `r shakhtest2$conf.int[2]`), `r cfig("shakhplot")`c), suggesting that reducing gene expression levels at a controlling CpG typically leads to increased gene expression levels, consistent with expectation [@Bird2002].

@@

Researchers are often confronted with the problem of making causal inferences using a statistical framework on observational data. In the epidemiological literature issues of measurement error in mediation analysis is relatively well explored [@Fewell2007]. Our analysis extends this to related methods such as CIT that are widely used in predominantly 'omic data. These methods are indeed liable to the same issues as standard mediation based analysis, and specifically we show that as measurement error in the exposure variable increases, CIT is likely to have reduced statistical power, and liable to infer the wrong direction of causality. We also demonstrate that, though unintuitive, increasing sample size does not resolve the issue, rather it leads to more extreme p-values for the model that predicts the wrong direction of causality.

@@

We demonstrate this new method by evaluating the causal relationships of `r nrow(qtl_orig)` known associations between DNA methylation and gene expression levels. We demonstrate that inferring the causal direction is heavily influenced by how much measurement error is present in the different assaying platforms. For example, if DNA methylation measures typically have higher measuremet error than gene expression measures then our analysis suggests that DNA methylation levels would be more often the causal factor in the association. Indeed, previous studies which have evaluated measurement error in these platforms do support this position [@Bose2014; @Bryant2011], though making strong conclusions for this analysis is difficult because measurement error is likely to be study specific.

Mediation based network approaches, that go beyond analyses of two variables, are very well established [@Lagani2015] and have a number of extensions that make them valuable tools, including for example network construction. But because they are predicated on the basic underlying principles of mediation they are liable to suffer from the same issues of measurement error. Recent advances in MR methodology, for example applying MR to genetical genomics [@Relton2012], multivariate MR [@Burgess2014a], network MR [@Burgess2015] and mediation through MR [@Varbo2015] may offer more robust alternatives for these more complicated problems.

In our simulations we focused on the simple case of a single instrument in a single sample setting, and assumed that pleiotropy (the influence of the instrument on the outcome through a mechanism other than the exposure) was not present. Recent method developments in MR [@Bowden2015; @Bowden2016] have focused on accounting for the issues that horizontal pleiotropy can introduce, but how they perform in the presence of measurement error remains to be explored. An important advantage that MR confers over most mediation based analysis is that it can be performed in two samples, which can considerably improve power and expand the scope of analysis. However, whether there is a substantive difference in two sample MR versus one sample MR in how measurement error has an effect is not yet fully understood. We have also assumed no measurement error in the genetic instrument, which is not unreasonable given the strict QC protocols that ensure high quality genotype data is available to most studies. We have restricted the scope to only exploring non-differential measurement error and avoided the complications incurred if measurement error in the exposure and outcome is correlated, however our analysis goes beyond many previous explorations of measurement error by assessing the impacts of both imprecision (noise) and linear transformations of the true variable on causal inference.
